NDNS multilevel LCA CW2CB2 or CW2CB3

### 2018-08-20 paper/Appendices/AppendixE.tex
abstract almost done, proof reading first round done

\begin{figure}[H]
%\vspace*{13cm}
\centering
\includegraphics{Figures/Diary1.png}
\decoRule
\caption[One day food diary example 6 am to 2 pm]{NATIONAL DIET AND NUTRITION SURVEY -- Food and Drink Diary Example, from 6 am to 2 pm.}
\label{fig:diary1}
\end{figure}
%\end{verbatim}

\begin{figure}[H]
%\vspace*{13cm}
\centering
\includegraphics{Figures/Diary2.png}
\decoRule
\caption[One day food diary example 2 pm to 6 am]{NATIONAL DIET AND NUTRITION SURVEY -- Food and Drink Diary Example, from 2 pm to 6 am.}
\label{fig:diary2}
\end{figure}
%\end{verbatim}

\begin{figure}[H]
%\vspace*{13cm}
\centering
\includegraphics{Figures/Diary3.png}
\decoRule
\caption[Food diary example of home made food recipes]{NATIONAL DIET AND NUTRITION SURVEY -- Food and Drink Diary Example, home made food recipes.}
\label{fig:diary3}
\end{figure}

>>>>

\begin{figure}[H]
%\vspace*{13cm}
\centering
\includegraphics{Figures/Diary1.png}
\decoRule
\caption[One day food diary example 6 am to 12 noon.]{NATIONAL DIET AND NUTRITION SURVEY -- Food and Drink Diary Example, from 6 am to 12 noon.}
\label{fig:diary1}
\end{figure}
%\end{verbatim}

\begin{figure}
%\vspace*{13cm}
\centering
\includegraphics{Figures/Diary2.png}
\decoRule
\caption[One day food diary example 12 noon to 5 pm]{NATIONAL DIET AND NUTRITION SURVEY -- Food and Drink Diary Example, from 12 noon to 5 pm.}
\label{fig:diary2}
\end{figure}
%\end{verbatim}

\begin{figure}
	%\vspace*{13cm}
	\centering
	\includegraphics{Figures/Diary03.png}
	\decoRule
	\caption[One day food diary example 5 pm to 6 am]{NATIONAL DIET AND NUTRITION SURVEY -- Food and Drink Diary Example, from 5 pm to 6 am.}
	\label{fig:diary2}
\end{figure}
%\end{verbatim}

\begin{figure}
%\vspace*{13cm}
\centering
\includegraphics{Figures/Diary3.png}
\decoRule
\caption[Food diary example of home made food recipes]{NATIONAL DIET AND NUTRITION SURVEY -- Food and Drink Diary Example, home made food recipes.}
\label{fig:diary3}
\end{figure}

### 2018-08-20 paper/Chapters/Chapter2.tex
abstract almost done, proof reading first round done

Time across a typical survey day was divided into 7 time slots in the dietary diary of NDNS RP: 6 am to 9 am, 9 am to 12 noon, 12 noon to 2 pm, 2 pm to 5 pm, 5 pm to 8 pm, 8 pm to 10 pm, and 10 pm to 6 am next morning. To produce a sequence of discrete responses regarding the carbohydrate intake we are interested, the energy consumption within each time slot over the four days of survey for each participant were calculated. The percentages of energy that contributed by carbohydrate within each time slot were then estimated. Since we planed to apply latent class analysis (LCA) in the current study, in which the observed indicators for latent classes must be categorical, we then dichotomised the responses according to the carbohydrate contribution to the energy intake at cut-off value of 50\%, i.e. if within a time slot where there is any energy intake occured, carbohydrate consumption was categorised into whether it's energy contribution was lower or higher/equal to 50\% of total energy intake within that time slot. Consequently, for each day of the recording, there were 7 data points generated by the diary, each data point included one of the following responses:

@@

Computer assisted personal interviews were conducted for the selected individuals by trained interviewers to collect background information on smoking habits (current, ex-smokers, and never), ethnicity (white, non-white), education level (lower than degree/degree or above level), living with a partner or not and other socio-demographic related variables. Participants also had their height, weight, blood pressure, waist circumferences (WC) measured by the nurses.

Specifically, blood pressure was measured in a sitting position using an automated, validated machine, the Omron HEM907, after a five minute rest. The means of second and third readings, taken at one minute intervals, were used in the current report. Hypertension was defined as with systolic blood pressure of 140 mmHg or above, and/or diastolic blood pressure of 90 mmHg or above, and/or taking any medication specifically to reduce blood pressure.

@@

\begin{equation}
P(U_{i1} = s_1, U_{i2} = s_2, \cdots, U_{ik} = s_K) = \sum_{t=t}^{T}P(C_i = t)\prod_{k = 1}^{K}P(U_{ik} = s_k | C_i = t)
\label{LCA}
\end{equation}\vspace{-0.7cm}

@@

\begin{itemize}
	\item $P(U_{i1} = s_1, U_{i2} = s_2, \cdots, U_{ik} = s_k)$ is the probability of observing a particular vector of responses;
	\item $P(C_i = t)$ is the probability that a randomly selected $i$th observation will be in class $t$;
	\item $P(U_{ik} = s_k | C_i = t)$ is the probability of a particular observed response pattern $U_{ik} = s_k$ conditional on membership in latent class $t$.
\end{itemize}

\textbf{Equation \ref{LCA}} indicates that responses for an observation to the measuring variables are independent of one another given its membership in latent class $t$. However, in the NDNS RP data set, the assumption of independent observations is violated. Each individual completed their dietary diary for four consecutive days, their diary recordings were later converted into four sequences of categorical responses reflecting the type of carbohydrate consumption at each time slot of the day. The four observed sequences (observed days) are nested in the participants and therefore are not independent. This nested data structure requires multilevel techniques. 
\vspace{-0.3cm}

@@

\begin{itemize}
	\item $\text{logit}{(x)} = \frac{\log(x)}{\log(1-x)}$;
	\item $P(C_{ij} = 1)$ as the probability that the randomly selected $i$th observation day of $j$th individual is a high carbohydrate eating day;
	\item $\beta_{0j}$ as the random intercept, for outcome $C_{ij} = 1$; 
	\item the random deviation of the individuals $u_{0j}$ are assumed be normally distributed (i.e. $u_{0j} \sim N(0, \sigma_{u_0}^2)$), the magnitude of the $u_{0j}$ variance ($\sigma_{u_0}^2$) indicates the influence of the individuals (level 2);
	\item $x_{ij}, w_j$ is the predictors for day level (weekdays or weekends) and individual level, such as age, and sex.
\end{itemize}

@@

If the day level LCA model (carbohydrate eating temporal pattern typologies) is best defined by $T (T \geqslant 2)$ latent classes, then $T-1$ random intercept will be specified by a two-level multinomial logistic regression model. Similar to the typical LCA models, the latent class variable in a MLCA is defined by multiple observed indicators (here is defined by the responses of eating carbohydrate within each time slots, throughout 4 consecutive days of survey period). Considering the latent class indicators are indicator variables ($U_{ijk}$), the MLCA model can be written as follows:\vspace{-1cm}

\begin{equation}
%\begin{aligned}
P(U_{ij1} = s_1, U_{ij2} = s_2, \cdots, U_{ijk} = s_{K}) = \sum_{t=1}^{T}P(G_{ij}=t)\prod_{k=1}^{K}P(U_{ijk} = s_k | G_{ij} = t)
%\end{aligned}
\label{MLCA}
\end{equation}
\vspace{-0.8cm}

@@

\begin{itemize}
	\item $ U_{ijk} $ represents the response of eating carbohydrate (one of the following: not eating any food, $< 50\%$ of the energy, or $\geqslant 50\%$ of the energy) on $i$th day of the survey ($i \in (1,2,3,4)$) in $j$th individual at the $k$th time slot of the day ($k \in (1, 2, 3, \cdots, 7)$);
	\item $G_{ij}$ denotes the latent class membership for $j$th individuals on $i$th day of the survey, the total number of day level latent class is $T$;
%	\item A specific latent class is referred to as $t$, and the total number of level 1 latent classes is denoted by $T$;
	\item $P(U_{ijk} = s_k|C_{ij} = t)$ is the probability of a specific response pattern, conditional on membership in latent class $t$.
\end{itemize}

@@

Since the parametric approach discussed above can be extremely computationally demanding \parencite{van2008using, vermunt2008latent}, an alternative approach is using a non-parametric MLCA \parencite{davidian2008growth}. In this approach, separate latent class models are specified for level 1 (observation days) and level 2 (individuals). Similar with the parametric MLCA approach, there are $T-1$ random intercepts, where $T$ is the number of level 1 latent classes. However, rather than assuming the random intercepts following a normal distribution, the non-parametric MLCA assumes a multinomial (discrete) distribution of the level 2 latent classes. This approach is less computationally demanding compared with the parametric approach. These level 2 (individual) latent classes reflect differences in the probability of belonging to a specific day level latent class, so that individuals that contain observation days with similar probabilities for the level 1 latent classes will be grouped together. The non-parametric MLCA model can be defined as follows: \vspace{-0.7cm}

@@

\begin{itemize}
	\item Firstly, we ignored the multilevel structure of the data and estimated a series of traditional LC models to determine the number of classes at the observational day-level;
	\item Next, a series of MLCA models were fitted to account for the multilevel structrure of the data. In these models, the number of day-level classes was based on the best fitting LCA model from the first step, and the LCA model at the individual level was estimated to identify the number of individual level latent classes;
	\item Thirdly, when number of individual level latent classes is defined based on the previous stage, day-level classes was modified (one class lower and one class higher than in the second step) to see the effect of changing level 1 classes and confirm the best fitting model.
\end{itemize}

The number of classes in either level 1 were determined by 1) the evaluation of model fit indices, including the Bayesian information criterion (BIC), and entropy which is a statistic that summarizes latent class probabilities where values near 1 indicate better latent class separation; 2) the Lo-Mendell-Rubin Likelihood Ratio Test (LMR-LRT) \parencite{lo2001testing, nylund2007deciding} which compares $q$ vs. $q-1$ classes models, where $q$ is the number of latent classes and most importantly, 3) pattern interpretability. In the steps of performing multilevel LCA, where LMR-LRT is not available, same rules of model fit indices and pattern interpretability were used to determine the optimal combination of latent classes in observation day level and individual level. MLCA models were fitted in Mplus 7.4 \parencite{muthen2005mplus}, the Mplus syntax and outputs are shown in \textbf{Appendix \ref{AppendixB}}. \vspace{-0.3cm}

@@

Person level point estimates and 95\% confidence intervals (CIs) were determined by applying individual, nurse visiting, and blood sample weights accordingly which account for the probability of participant selection and the clustered survey design. Descriptive statistics for sample characteristics are presented as weighted means (95\% CI) or weighted percentages (95\%CI). After examining the distribution of the data, the following variables were log-transformed to improve normality: fasting blood glucose, A1C, TC, LDL, HDL, TG, and average physical activity duration per day. Weighted geometric means (95\% CI) were used for all log-transformed variables.

@@

\begin{enumerate}
	\item The crude association between the carbohydrate eating groups and the outcomes was first examined. 
	\item Potential confounders of the association between carbohydrate eating groups (exposure) and the outcomes were selected depending on the descriptive statisitical analyses conducted above, i.e. those are associated with both the exposure and the outcome and also not on the causal pathway were selected as potential confounders. Those are strongly related with the outcomes but may not associated with carbohydrate eating groups may reduce the standard errors and so improve the precisions are also considered in the linear regression models. 
	\item Confounding and/or interaction effect from each of the potential factors were checked one by one. Interaction effect were tested using the adjusted Wald test testing whether the regression coefficients of the interaction terms are simultaneously equal to zero.  
	\item A preliminary model that includes all of the variables suggested to be confounders in the previous step was established. 
	\item The remaining variables were added to the preliminary model one by one to see if any of them may be a confounder in condition of the presence of the other covariates. 
	\item For logistic regression models (hypertension) under the survey data, goodness-of-fit was assessed using the adapted \code{svylogitgof} command in Stata \parencite{archer2006goodness}. Other diagnostics for regular logistic regression models, such as estimating the pseudo-R\textsuperscript{2},  AIC or BIC, checking the standardized Pearson residuals, or covariate pattern residuals are currently not available for weighted survey data. 
	\item For linear regression models (WC, BMI), assumption of independent observations is violated as soon as we weighted the sample. General checking such as QQ plots of the residuals (normality), plotting the residuals against fitted values (constant variance) are not available as well. Outliers, leverage, and Cook's distance cannot be check either, however, participants with extreme weightings (if exist) were checked by removing them and refit the models as a sensitivity analysis.
	\item Since under survey design data, the sampling-weighted least squares are not maximum likelihood, it would not be possible to compare models using likelihood ratio test. Instead, adjusted Wald tests with $p < 0.05$ were used as criteria for variable inclusion in the final model. Another Stata command \code{linktest} was also used to  decide whether quadratic and cubic terms of continuous variables were necessary in improving the fitting of model \parencite{pregibon1980goodness}. 
\end{enumerate}

>>>>

Time across a typical survey day was divided into 7 time slots in the dietary diary of NDNS RP: 6 am to 9 am, 9 am to 12 noon, 12 noon to 2 pm, 2 pm to 5 pm, 5 pm to 8 pm, 8 pm to 10 pm, and 10 pm to 6 am next morning. To produce a sequence of discrete responses regarding the carbohydrate intake we are interested, the energy consumption within each time slot over the four days of survey for each participant were calculated. The percentages of energy that contributed by carbohydrate within each time slot were then estimated. Since we planned to apply latent class analysis (LCA) in the current study, in which the observed indicators for latent classes must be categorical, the responses were then dichotomised according to the carbohydrate contribution to the energy intake at cut-off value of 50\%, i.e. if within a time slot where there is any energy intake occured, carbohydrate consumption was categorised into whether it's energy contribution was lower or higher/equal to 50\% of total energy intake within that time slot. Consequently, for each day of the recording, there were 7 data points generated by the diary, each data point included one of the following responses:

@@

Computer assisted personal interviews were conducted for the selected individuals by trained interviewers to collect background information on smoking habits (current, ex-smokers, and never), ethnicity (white, non-white), education level (lower than degree/degree or above level), living with a partner or not and other socio-demographic variables. Participants also had their height, weight, blood pressure, waist circumferences (WC) measured by the nurses.

Specifically, blood pressure was measured in a sitting position using an automated, validated machine, the Omron HEM907, after a five-minute rest. The means of second and third readings, taken at one minute intervals, were used in the current report. Hypertension was defined as with systolic blood pressure of 140 mmHg or above, and/or diastolic blood pressure of 90 mmHg or above, and/or taking any medication specifically to reduce blood pressure.

@@

\begin{equation}
P(U_{i1} = s_1, U_{i2} = s_2, \cdots, U_{ik} = s_k) = \sum_{t=t}^{T}P(C_i = t)\prod_{k = 1}^{K}P(U_{ik} = s_k | C_i = t)
\label{LCA}
\end{equation}\vspace{-0.7cm}

@@

\begin{itemize}
	\item $P(U_{i1} = s_1, U_{i2} = s_2, \cdots, U_{ik} = s_k)$ is the probability of observing a particular vector of responses for $i$th observation;
	\item $P(C_i = t)$ is the probability that a randomly selected $i$th observation will be in class $t$;
	\item $P(\mathbf{U_{ik}} = \mathbf{s_k} | C_i = t)$ is the probability of a particular observed response pattern $\mathbf{U_{ik}} = \mathbf{s_k}$ conditional on membership in latent class $t$.
\end{itemize}

\textbf{Equation \ref{LCA}} indicates that responses for an observation to the measuring variables are independent of one another given its membership in latent class $t$. However, in the NDNS RP data set, the assumption of independent observations is violated. Each individual completed their dietary diary for three/four consecutive days, their diary recordings were later converted into three/four sequences of categorical responses reflecting the type of carbohydrate consumption at each time slot of the day. The observed sequences (observed days) are nested within the participants and therefore are not independent. This nested data structure requires multilevel techniques. 
\vspace{-0.3cm}

@@

\begin{itemize}
	\item $\text{logit}{(x)} = \log{\frac{x}{1-x}}$, where $x \in [0,1]$;
	\item $P(C_{ij} = 1)$ as the probability that the randomly selected $i$th observation day of $j$th individual is a high carbohydrate eating day;
	\item $\beta_{0j}$ as the random intercept, for outcome $C_{ij} = 1$; 
	\item the random deviation of the individuals $u_{0j}$ are assumed be normally distributed (i.e. $u_{0j} \sim N(0, \sigma_{u_0}^2)$), the magnitude of the $u_{0j}$ variance ($\sigma_{u_0}^2$) indicates the influence of the individuals (level 2);
	\item $x_{ij}, w_j$ is the predictors for day level (weekdays or weekends) and individual level, such as age, and sex.
\end{itemize}

@@

If the day level LCA model (carbohydrate eating temporal pattern typologies) is best defined by $T (T \geqslant 2)$ latent classes, then $T-1$ random intercept will be specified by a two-level multinomial logistic regression model. Similar to the typical LCA models, the latent class variable in a MLCA is defined by multiple observed indicators (here is defined by the responses of eating carbohydrate within each time slots, throughout three/four consecutive days of their survey period). Considering the latent class indicators are indicator variables ($U_{ijk}$), the MLCA model can be written as follows:\vspace{-1cm}

\begin{equation}
%\begin{aligned}
P(U_{ij1} = s_1, U_{ij2} = s_2, \cdots, U_{ijk} = s_{k}) = \sum_{t=1}^{T}P(G_{ij}=t)\prod_{k=1}^{K}P(U_{ijk} = s_k | G_{ij} = t)
%\end{aligned}
\label{MLCA}
\end{equation}
\vspace{-0.8cm}

@@

\begin{itemize}
	\item $ U_{ijk} $ represents the response of carbohydrate eating (one of the following: not eating any food, $< 50\%$ of the energy, or $\geqslant 50\%$ of the energy) on $i$th day of the survey ($i \in (1,2,3,4)$) in $j$th individual at the $k$th time slot of the day ($k \in (1, 2, 3, \cdots, 7)$);
	\item $G_{ij}$ denotes the latent class membership for $j$th individuals on $i$th day of the survey, the total number of day level latent class is $T$;
%	\item A specific latent class is referred to as $t$, and the total number of level 1 latent classes is denoted by $T$;
	\item $P(U_{ijk} = s_k|C_{ij} = t)$ is the probability of a specific response pattern, conditional on membership in latent class $t$.
\end{itemize}

@@

Since the parametric approach discussed above can be extremely computationally demanding \parencite{van2008using, vermunt2008latent}, an alternative approach is using a non-parametric MLCA \parencite{davidian2008growth}. In this approach, separate latent class models are specified for level 1 (observation days) and level 2 (individuals). Similar with the parametric MLCA approach, there are $T-1$ random intercepts, where $T$ is the number of level 1 latent classes. However, rather than assuming the random intercepts following a normal distribution, the non-parametric MLCA assumes a multinomial (discrete) distribution of the level 2 latent classes. This approach is less computationally demanding compared with the parametric approach. These level 2 (individual) latent classes reflect differences in the probability of belonging to a specific day level latent class, so that individuals that have observation days with similar probabilities for the level 1 latent classes will be grouped together. The non-parametric MLCA model can be defined as follows:

@@

\begin{itemize}
	\item Firstly, we ignored the multilevel structure of the data and estimated a series of traditional LC models to determine the number of classes at the observational day-level;
	\item Next, a series of MLCA models were fitted to account for the multilevel structrure of the data. In these models, the number of day-level classes was based on the best fitting LCA model from the first step, and the LCA model at the individual level was estimated to identify the number of individual level latent classes;
	\item Thirdly, when number of individual level latent classes is defined based on the previous stage, number of day-level classes was modified (one class lower and one class higher than in the second step) to see the effect of changing level 1 classes and confirm the best fitting model.
\end{itemize}

The number of classes in level 1 was determined by 1) the evaluation of model fit indices, including the Bayesian information criterion (BIC), and entropy which is a statistic that summarizes latent class probabilities where values near 1 indicate better latent class separation; 2) the Lo-Mendell-Rubin Likelihood Ratio Test (LMR-LRT) \parencite{lo2001testing, nylund2007deciding} which compares $q$ vs. $q-1$ classes models, where $q$ is the number of latent classes and most importantly, 3) pattern interpretability. In the steps of performing multilevel LCA, where LMR-LRT is not available, same rules of model fit indices and pattern interpretability were used to determine the optimal combination of latent classes in observation day level and individual level. MLCA models were fitted in Mplus 7.4 \parencite{muthen2005mplus}, the Mplus syntax and outputs are shown in \textbf{Appendix \ref{AppendixB}}. \vspace{-0.3cm}

@@

Person level point estimates and 95\% confidence intervals (CIs) were determined by applying individual, nurse visiting, and blood sample weights accordingly which account for the probability of participant selection and the clustered survey design. Descriptive statistics for sample characteristics were presented as weighted means (95\% CI) or weighted percentages (95\%CI). After examining the distribution of the data, the following variables were log-transformed to improve normality: fasting blood glucose, A1C, TC, LDL, HDL, TG, and average physical activity duration per day. Weighted geometric means (95\% CI) were used for all log-transformed variables.

@@

\begin{enumerate}
	\item The crude association between the carbohydrate eating groups and the outcomes was first examined. 
	\item Potential confounders of the association between carbohydrate eating groups (exposure) and the outcomes were selected depending on the descriptive statisitical analyses conducted above, i.e. those are associated with both the exposure and the outcome and also not on the causal pathway were selected as potential confounders. Those are strongly related with the outcomes but may not associated with carbohydrate eating groups may reduce the standard errors and so improve the precisions are also considered in the linear regression models. 
	\item Confounding and/or interaction effect from each of the potential factors were checked one by one. Interaction effect were tested using the adjusted Wald test testing whether the regression coefficients of the interaction terms are simultaneously equal to zero.  
	\item A preliminary model that includes all of the variables suggested to be confounders in the previous step was established. 
	\item The remaining variables were added to the preliminary model one by one to see if any of them may be a confounder in condition of the presence of the other covariates. 
	\item For logistic regression models (hypertension) under the survey data, goodness-of-fit was assessed using the adapted \code{svylogitgof} command in Stata \parencite{archer2006goodness}. Other diagnostics for regular logistic regression models, such as estimating the pseudo-R\textsuperscript{2},  AIC or BIC, checking the standardized Pearson residuals, or covariate pattern residuals are currently not available for weighted survey data. 
	\item For linear regression models (WC, BMI), assumption of independent observations is violated as soon as we weighted the sample. General checking such as QQ plots of the residuals (normality), plotting the residuals against fitted values (constant variance) are not available as well. Outliers, leverage, and Cook's distance cannot be checked either, however, participants with extreme weightings (if exist) were checked by removing them and refit the models as a sensitivity analysis.
	\item Since under survey design data, the sampling-weighted least squares are not maximum likelihood, it would not be possible to compare models using likelihood ratio test. Instead, adjusted Wald tests with $p < 0.05$ were used as criteria for variable inclusion in the final model. Another Stata command \code{linktest} was also used to  decide whether quadratic and cubic terms of continuous variables were necessary in improving the fitting of model \parencite{pregibon1980goodness}. 
\end{enumerate}
